---
title: "What is your blogpost title? Add it here??"
subtitle: "Assignment 4 ETC5512"
author: "Who are you?? Add your name here"
format: html
editor: visual
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data and Documenting

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(here)
library(readr)

```

```{r Reading-Data}
underemp_total <- read_csv("data/Chart 1_ Underemployed workers, Original.csv", skip= 1)

underemp_gender <- read_csv("data/Chart 5_ Underemployed workers, by Sex, Original.csv")

underutilisation <- read_csv("data/Underemployment rate and underutilisation rate.csv")

unemployment <- read_csv ("data/Unemployment rate.csv")

readLines(
  "data/Chart 1_ Underemployed workers, Original.csv",
  n = 3
)

```


```{r Cleaning-Dataformat}
#total underemployment 

underemp_total <- underemp_total %>%
  separate(
    1,
    into = c(
      "month",
      "hours_reduced",
      "prefers_more_hours",
      "total_underemployed"
    ),
    sep = ","
  )

#total underemployment by Gender

underemp_gender <- underemp_gender |> 
  separate(
    1,          #Use the first column in the dataframe.
    into = c(
      "month",
      "male_prefers_more_hours",
      "male_hours_reduced",
      "female_prefers_more_hours",
      "female_hours_reduced"
    ),
    sep = ","
  )

#unemployment rate

unemployment <- unemployment |> 
  separate(
    1,
    into = c(
      "month",
      "unemployment_rate_trend",
      "unemployment_rate_sa"
    ),
    sep = ","
  )

#Rate of Underemployment and Underutilisation

underutilisation <- underutilisation |> 
  separate(
    1,
    into = c(
      "month",
      "underemployment_rate_trend",
      "underemployment_rate_sa",
      "underutilisation_rate_trend",
      "underutilisation_rate_sa"
    ),
    sep = ","
  )

```


```{r Coverting-DataType}
underemp_total <- underemp_total |>
  mutate(
    hours_reduced = parse_number(hours_reduced),
    prefers_more_hours = parse_number(prefers_more_hours),
    total_underemployed = parse_number(total_underemployed)
  )

underemp_gender <- underemp_gender %>%
  mutate(
    male_prefers_more_hours = as.numeric(male_prefers_more_hours),
    male_hours_reduced = as.numeric(male_hours_reduced),
    female_prefers_more_hours = as.numeric(female_prefers_more_hours),
    female_hours_reduced = as.numeric(female_hours_reduced)
  )

underutilisation <- underutilisation %>%
  mutate(
    underemployment_rate_trend =
      as.numeric(underemployment_rate_trend),

    underemployment_rate_sa =
      as.numeric(underemployment_rate_sa),

    underutilisation_rate_trend =
      as.numeric(underutilisation_rate_trend),

    underutilisation_rate_sa =
      as.numeric(underutilisation_rate_sa)
  )

unemployment <- unemployment %>%
  mutate(
    unemployment_rate_trend =
      as.numeric(unemployment_rate_trend),

    unemployment_rate_sa =
      as.numeric(unemployment_rate_sa)
  )


```



### What's in this section

Here is where you should provide the more comprehensive details about your data and your processing steps.

These are the data details that an "average" blog post reader is not concerned with (Task 1).

### Remember

Please mention any additional files that you want the markers to review, eg. read me, meta data etc.

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### What's in this section

Here is where you should write your blogpost! (Task 2)

### Blogpost Structure

There is no strict structure for you to follow. However, here is a skeleton structure you may find helpful.

1.  Title (is set at the top of this document)
2.  Motivation\
3.  Data\
4.  Analysis\
5.  Conclusions\
6.  References

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.
:::
