---
title: "Understanding Gender Differences in Underemployment contributing to Underutilisation of Labor in Australia"
subtitle: "Assignment 4 | ETC5512 Wild-Caught Data"
author: "Somya Verma"
format:
  html:
    theme: superhero
toc: true
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data and Documenting

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(here)
library(readr)

```

```{r Reading-Data}
underemp_total <- read_csv("data/Chart 1_ Underemployed workers, Original.csv", skip= 1)

underemp_gender <- read_csv("data/Chart 5_ Underemployed workers, by Sex, Original.csv")

underutilisation <- read_csv("data/Underemployment rate and underutilisation rate.csv")

unemployment <- read_csv ("data/Unemployment rate.csv")

```


```{r Cleaning-Dataformat}
#total underemployment 

underemp_total <- underemp_total |>
  rename(
    month = month,
    hours_reduced = "Worked less than usual hours for economic reasons ('000) ('000)",
    prefers_more = "Prefers more hours than usual ('000) ('000)",
    total_underemp = "Total ('000) ('000)"
  )

#total underemployment by Gender

underemp_gender <- underemp_gender |> 
  separate(
    1,          #Use the first column in the dataframe.
    into = c(
      "month",
      "male_prefers_more_hours",
      "male_hours_reduced",
      "female_prefers_more_hours",
      "female_hours_reduced"
    ),
    sep = ","
  )

#unemployment rate

unemployment <- unemployment |> 
  separate(
    1,
    into = c(
      "month",
      "unemployment_rate_trend",
      "unemployment_rate_sa"
    ),
    sep = ","
  )

#Rate of Underemployment and Underutilisation

underutilisation <- underutilisation |> 
  separate(
    1,
    into = c(
      "month",
      "underemployment_rate_trend",
      "underemployment_rate_sa",
      "underutilisation_rate_trend",
      "underutilisation_rate_sa"
    ),
    sep = ","
  )

underutilisation <- underutilisation |>
  filter(month != "")

underutilisation <- underutilisation |>
  slice(-1)

```


```{r Coverting-DataType}

underemp_total <- underemp_total |>
  mutate(
    hours_reduced = as.numeric(hours_reduced),
    prefers_more = as.numeric(prefers_more),
    total_underemp = as.numeric(total_underemp)
  )


underemp_gender <- underemp_gender |>
  mutate(
    male_prefers_more_hours =
      as.numeric(male_prefers_more_hours),

    male_hours_reduced =
      as.numeric(male_hours_reduced),

    female_prefers_more_hours =
      as.numeric(female_prefers_more_hours),

    female_hours_reduced =
      as.numeric(female_hours_reduced)
  )

underutilisation <- underutilisation |>
  slice(-1)
underutilisation <- underutilisation |>
  mutate(
    underemployment_rate_trend =
      as.numeric(underemployment_rate_trend),

    underemployment_rate_sa =
      as.numeric(underemployment_rate_sa),

    underutilisation_rate_trend =
      as.numeric(underutilisation_rate_trend),

    underutilisation_rate_sa =
      as.numeric(underutilisation_rate_sa)
  )

unemployment <- unemployment |>
  slice(-1)
unemployment <- unemployment |>
  mutate(
    unemployment_rate_trend = as.numeric(unemployment_rate_trend),
    unemployment_rate_sa = as.numeric(unemployment_rate_sa)
  )
glimpse(underutilisation)

underemp_gender <- underemp_gender |>
  filter(!is.na(month))
underutilisation <- underutilisation |>
  filter(!is.na(month))

unemployment <- unemployment |>
  filter(!is.na(month))

```

```{r}
underemp_gender <- underemp_gender |>
  filter(month != "month")

head(underemp_gender$month)
head(underutilisation$month)
head(unemployment$month)


underemp_total <- underemp_total |>
  mutate(
    month = my(month)
  )

underemp_gender <- underemp_gender |>
  mutate(
    month = my(month)
  )

underutilisation <- underutilisation |>
  mutate(
    month = my(month)
  )

unemployment <- unemployment |>
  mutate(
    month = my(month)
  )

underemp_gender |>
  count(month) |>
  filter(n > 1)

underemp_gender <- underemp_gender |>
  filter(!is.na(month))

underemp_gender |>
  filter(is.na(month))

underutilisation <- underutilisation |>
  filter(!is.na(month))

unemployment <- unemployment |>
  filter(!is.na(month))

underutilisation |>
  count(month) |>
  filter(n > 1)

unemployment |>
  count(month) |>
  filter(n > 1)

underemp_total |>
  count(month) |>
  filter(n > 1)

labour_data <- underemp_gender |>
  left_join(
    underutilisation,
    by = "month"
  ) |>
  left_join(
    unemployment,
    by = "month"
  ) |>
  filter(
    month >= as.Date("2016-06-01"),
    month <= as.Date("2025-12-31")
  )


```


### Data Overview

Four datasets were obtained from the Australian Bureau of Statistics (ABS) Labour Force publications to investigate the relationship between gender differences in underemployment and labour underutilisation in Australia. The datasets included total underemployed workers, underemployed workers by sex, underemployment and underutilisation rates, and unemployment rates.

The data used in this analysis were obtained from the Australian Bureau of Statistics (ABS) [Labour Force Survey](https://www.abs.gov.au/methodologies/labour-force-australia-methodology/apr-2026), a monthly survey of approximately 50,000 individuals designed to measure employment and unemployment outcomes across Australia. The ABS publishes labour market indicators as original, seasonally adjusted, and trend series. For this analysis, seasonally adjusted estimates were used where available because they remove recurring seasonal patterns and allow underlying labour market movements to be compared more effectively over time.

The [ABS](https://www.abs.gov.au/methodologies/labour-force-australia-methodology/apr-2026) notes that labour market conditions experienced substantial disruption during the COVID-19 period (April 2020 to March 2022), resulting in increased volatility and several trend breaks. Consequently, caution is exercised when interpreting labour market movements during this period.

#### Data Processing

**Data Import**

* Imported all ABS datasets into R using the `read_csv()` function from the readr package.
Excluded the title row from the total underemployment dataset using skip = 1.

**Data Cleaning and Transformation**

* Renamed variables to shorter and more interpretable names for ease of analysis.
Used separate() from the tidyr package to split comma-separated text columns into individual variables.

* Converted labour market measures from `character (chr)` to `numeric (dbl)` data types using `as.numeric()`.

* Removed duplicated header rows from the underutilisation and unemployment datasets using `slice(-1)` to prevent conversion errors and missing values.

**Date Processing**

* Converted the month variable from the original ABS format (e.g., Apr-16) to a date format using `my()` from the `lubridate package`.

* Standardised the time variable across all datasets to support time-series analysis and dataset integration.

**Final Dataset Preparation**

* Ensured all datasets contained consistent date variables and correctly formatted numeric measures.

* Prepared the datasets for merging and analysis of underemployment, unemployment, and labour underutilisation trends. Derived new variables. 

* Focused the analysis on gender differences in underemployment in Australia between 2016 and 2025.


### Remember

Please mention any additional files that you want the markers to review, eg. read me, meta data etc.

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

```{r Deriving_Variables}

underemp_gender <- underemp_gender |>
  mutate(
    Total_male_underemp =
      male_prefers_more_hours +
      male_hours_reduced,
    Total_female_underemp =
      female_prefers_more_hours +
      female_hours_reduced
  )

#Creating a Gender Gap

underemp_gender <- underemp_gender |>
  mutate(
    gender_gap =
      Total_female_underemp - Total_male_underemp
  )


```



### What's in this section

Here is where you should write your blogpost! (Task 2)

### Blogpost Structure

There is no strict structure for you to follow. However, here is a skeleton structure you may find helpful.

1.  Title (is set at the top of this document)
2.  Motivation\
3.  Data\
4.  Analysis\
5.  Conclusions\
6.  References

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.
:::
