---
title: "Does Gender Differences in Underemployment contribute to Underutilisation of Labour in Australia?"
subtitle: "Assignment 4 | ETC5512 Wild-Caught Data"
author: "Somya Verma"
format:
  html:
    theme: superhero
toc: true
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data and Documenting

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(here)
library(readr)
```

```{r Reading-Data}
underemp_total <- read_csv("data/Chart 1_ Underemployed workers, Original.csv", skip= 1)

underemp_gender <- read_csv("data/Chart 5_ Underemployed workers, by Sex, Original.csv")

underutilisation <- read_csv("data/Underemployment rate and underutilisation rate.csv")

unemployment <- read_csv ("data/Unemployment rate.csv")

```


```{r Cleaning-Dataformat}
#total underemployment 

underemp_total <- underemp_total |>
  rename(
    month = month,
    hours_reduced = "Worked less than usual hours for economic reasons ('000) ('000)",
    prefers_more = "Prefers more hours than usual ('000) ('000)",
    total_underemp = "Total ('000) ('000)"
  )

#total underemployment by Gender

underemp_gender <- underemp_gender |> 
  separate(
    1,          #Use the first column in the dataframe.
    into = c(
      "month",
      "male_prefers_more_hours",
      "male_hours_reduced",
      "female_prefers_more_hours",
      "female_hours_reduced"
    ),
    sep = ","
  )

#unemployment rate

unemployment <- unemployment |> 
  separate(
    1,
    into = c(
      "month",
      "unemployment_rate_trend",
      "unemployment_rate_sa"
    ),
    sep = ","
  )

#Rate of Underemployment and Underutilisation

underutilisation <- underutilisation |> 
  separate(
    1,
    into = c(
      "month",
      "underemployment_rate_trend",
      "underemployment_rate_sa",
      "underutilisation_rate_trend",
      "underutilisation_rate_sa"
    ),
    sep = ","
  )

underutilisation <- underutilisation |>
  filter(month != "")

underutilisation <- underutilisation |>
  slice(-1)

```


```{r Coverting-DataType}

underemp_total <- underemp_total |>
  mutate(
    hours_reduced = as.numeric(hours_reduced),
    prefers_more = as.numeric(prefers_more),
    total_underemp = as.numeric(total_underemp)
  )


underemp_gender <- underemp_gender |>
  mutate(
    male_prefers_more_hours =
      as.numeric(male_prefers_more_hours),

    male_hours_reduced =
      as.numeric(male_hours_reduced),

    female_prefers_more_hours =
      as.numeric(female_prefers_more_hours),

    female_hours_reduced =
      as.numeric(female_hours_reduced)
  )

underutilisation <- underutilisation |>
  slice(-1)
underutilisation <- underutilisation |>
  mutate(
    underemployment_rate_trend =
      as.numeric(underemployment_rate_trend),

    underemployment_rate_sa =
      as.numeric(underemployment_rate_sa),

    underutilisation_rate_trend =
      as.numeric(underutilisation_rate_trend),

    underutilisation_rate_sa =
      as.numeric(underutilisation_rate_sa)
  )

unemployment <- unemployment |>
  slice(-1)
unemployment <- unemployment |>
  mutate(
    unemployment_rate_trend = as.numeric(unemployment_rate_trend),
    unemployment_rate_sa = as.numeric(unemployment_rate_sa)
  )
underemp_gender <- underemp_gender |>
  filter(!is.na(month))
underutilisation <- underutilisation |>
  filter(!is.na(month))

unemployment <- unemployment |>
  filter(!is.na(month))

```

```{r}
#| echo: false
#| results: false

underemp_gender <- underemp_gender |>
  filter(month != "month")

head(underemp_gender$month)
head(underutilisation$month)
head(unemployment$month)


underemp_total <- underemp_total |>
  mutate(
    month = my(month)
  )

underemp_gender <- underemp_gender |>
  mutate(
    month = my(month)
  )

underutilisation <- underutilisation |>
  mutate(
    month = my(month)
  )

unemployment <- unemployment |>
  mutate(
    month = my(month)
  )

underemp_gender |>
  count(month) |>
  filter(n > 1)

underemp_gender <- underemp_gender |>
  filter(!is.na(month))

underemp_gender |>
  filter(is.na(month))

underutilisation <- underutilisation |>
  filter(!is.na(month))

unemployment <- unemployment |>
  filter(!is.na(month))

underutilisation |>
  count(month) |>
  filter(n > 1)

unemployment |>
  count(month) |>
  filter(n > 1)

underemp_total |>
  count(month) |>
  filter(n > 1)

labour_data <- underemp_gender |>
  left_join(
    underutilisation,
    by = "month"
  ) |>
  left_join(
    unemployment,
    by = "month"
  ) |>
  filter(
    month >= as.Date("2016-06-01"),
    month <= as.Date("2025-12-31")
  )


```


### Data Overview

This blog investigates: Does gender difference in underemployment contribute to labour underutilisation in Australia? To answer this, three sub-questions guide the analysis:

* Is underemployment meaningfully linked to overall labour underutilisation in Australia between 2016 and 2025?
* Do males and females contribute differently to total underemployment over time?
* Is the gender gap in underemployment associated with changes in the labour underutilisation rate?

Why this question? Economic disruptions rarely affect all groups equally. Certain sectors absorb more shock than others, and because men and women are not distributed evenly across industries, labour market stress can fall differently across genders.This raises a genuine question: when underemployment surged in Australia, did one gender bear a disproportionate share of that burden. Did that imbalance drive overall labour underutilisation? 


**Source and Suitability** 

All data were obtained from the Australian Bureau of Statistics (ABS) Labour Force Survey which is a [Monthly household survey](https://www.abs.gov.au/methodologies/labour-force-australia-methodology/apr-2026) of approximately 50,000 Australians and 24,000 dwellings conducted to measure employment and unemployment outcomes. As a large-scale probability sample survey, it is the most authoritative source of labour market data in Australia and is widely used by government, researchers, and policymakers. The survey directly measures underemployment, underutilisation, and unemployment, making it well suited to answering the research question.

Four datasets were used:

* Total underemployed workers (original series)
* Underemployed workers by sex (original series)
* Underemployment and underutilisation rates (trend and seasonally adjusted)
* Unemployment rates (trend and seasonally adjusted)

Data Type

This is sample survey data — not a census or experiment. The ABS Labour Force Survey uses a rotating sample design, meaning results are estimates with associated sampling error rather than exact population counts.

Data Licence

All ABS data are published under the Creative Commons Attribution 4.0 International licence (CC BY 4.0), permitting free use, sharing, and adaptation with attribution.

Data Limitations

* As sample-based estimates, all figures carry sampling error and should not be interpreted as exact counts.
* The ABS notes that the COVID-19 period (April 2020 – March 2022) caused significant volatility and trend breaks in the series, meaning comparisons across this period require caution.
* The gender breakdown is limited to male and female categories only, reflecting the binary classification used in ABS data collection during this period. Non-binary and gender-diverse individuals are not captured.
* Original series data were used for underemployment counts, which means seasonal effects are not removed and may influence month-to-month comparisons.

Ethical Considerations

* The ABS Labour Force Survey is conducted under the Census and Statistics Act 1905, which legally protects the confidentiality of individual responses. Only aggregate, anonymised statistics are published — no individual-level data are used in this analysis.

* No personally identifiable information is present in any dataset used, meaning there are no privacy risks associated with this analysis.

* The binary gender classification in the data is a limitation worth acknowledging — findings described as "male" and "female" reflect ABS reporting categories and may not fully represent the diversity of gender identities in the Australian workforce.


### Downloading the Data

All datasets were downloaded directly from the ABS Labour Force Survey publications. To reproduce this:

1. Go to the ABS Labour Force Survey page: [Labour Force Survey](https://www.abs.gov.au/methodologies/labour-force-australia-methodology/apr-2026)
2. Navigate to the **Data downloads** section of the most recent release.
3. Download the following Excel tables and export the relevant sheets as `.csv` files:
   - **Table 1** → Total underemployed workers (Original series) → saved as `Chart 1_ Underemployed workers, Original.csv`
   - **Table 5** → Underemployed workers by Sex (Original series) → saved as `Chart 5_ Underemployed workers, by Sex, Original.csv`
   - Underemployment and underutilisation rates → saved as `Underemployment rate and underutilisation rate.csv`
   - Unemployment rates → saved as `Unemployment rate.csv`
4. Place all raw `.csv` files into a `data/` folder within the project directory before running any code.

### Processing Steps

**Step 1 — Import**

All four datasets were imported using `read_csv()` from the `readr` package. The total underemployment dataset required `skip = 1` to bypass the title row. The remaining three datasets were read as single comma-separated text columns requiring further restructuring.

**Step 2 — Restructuring**

- The total underemployment dataset was renamed using `rename()` to give variables shorter, interpretable names (`hours_reduced`, `prefers_more`, `total_underemp`).
- The gender, underutilisation, and unemployment datasets each arrived as a single column of comma-separated text. `separate()` from `tidyr` was used to split each into individual named variables.

**Step 3 — Removing Duplicate Headers**

The underutilisation and unemployment datasets contained a repeated header row after separation. This was removed using `slice(-1)`. The gender dataset contained a row where `month == "month"` which was removed using `filter()`.

**Step 4 — Type Conversion**

All labour market measure columns were converted from `character (chr)` to `numeric (dbl)` using `as.numeric()`. Rows that could not be converted became `NA` and were subsequently removed using `filter(!is.na(month))`.

**Step 5 — Date Formatting**

The `month` variable was converted from ABS format (e.g., `Apr-16`) to a proper R date object using `my()` from the `lubridate` package, applied consistently across all four datasets.

**Step 6 — Duplicate Checks**

`count(month)` followed by `filter(n > 1)` was run on each dataset to confirm no duplicated monthly observations existed before merging.

**Step 7 — Merging**

The four datasets were merged into a single working dataset (`labour_data`) using two successive `left_join()` operations on the `month` variable, anchored to the gender dataset. The final dataset was filtered to cover **June 2016 to December 2025**.

**Step 8 — Derived Variables**

The following variables were derived within `labour_data` using `mutate()`:

| Variable | Formula |
|---|---|
| `Tot_male_underemp` | `male_prefers_more_hours + male_hours_reduced` |
| `Tot_female_underemp` | `female_prefers_more_hours + female_hours_reduced` |
| `total_underemp` | `Tot_male_underemp + Tot_female_underemp` |
| `male_share` | `Tot_male_underemp / total_underemp × 100` |
| `female_share` | `Tot_female_underemp / total_underemp × 100` |
| `share_gap` | `male_share − female_share` |



### Remember

Kindly look into the ReadME file for further information of the Dataset and Question approach. 

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### Let us move Beyond Unemployment Rate!  

The unemployment rate is often used to assess labour market performance, but it does not capture all forms of unused labour capacity. Many employed individuals would prefer to work additional hours, making underemployment an important component of labour underutilisation. This blog examines whether gender differences in underemployment contributed to labour underutilisation in Australia between 2016 and 2025 by analysing trends in underemployment, labour underutilisation, and the relative contributions of males and females to underemployed workers.

According to **Australian Bureau of Statistics ** [ABS](https://www.abs.gov.au/statistics/detailed-methodology-information/concepts-sources-methods/labour-statistics-concepts-sources-and-methods/2023/concepts-and-sources/underutilised-labour) labour underutilisation measures the extent to which available labour resources are not fully used in the economy. It combines unemployment and underemployment, providing a broader view of labour market spare capacity than unemployment alone. Underemployment refers to employed individuals who would like to work more hours and are available to do so, or full-time workers whose hours have been reduced for economic reasons.

Using ABS data from 2016 to 2025, this study examines whether differences in male and female contributions to underemployment are associated with changes in labour underutilisation over time.

### Do Underemployment and Underutilisation make sense together?

@fig-underutilisation shows that underemployment and labour underutilisation move almost hand in hand. When underemployment rises, underutilisation rises too, and when it falls, underutilisation follows. This suggests that underemployment is a key contributor to overall labour underutilisation in Australia.

The most noticeable spike occurs during COVID-19 in 2020, when both measures increased dramatically as many workers experienced reduced hours and labour market disruptions. While underutilisation remains consistently higher because it also includes unemployed workers, the similar patterns of the two lines indicate that underemployment plays a major role in explaining unused labour capacity. So are these two variables an item? YES! 

If underemployment contributes so strongly to labour underutilisation, the next question is whether males and females contribute differently to underemployment and whether these differences matter for overall labour market outcomes.

```{r Establishing-Relationship}
#| label: fig-underutilisation
#| fig-cap: "Underemployment and labour underutilisation rates in Australia, 2016–2025."
rates_long <- labour_data |>
  select(
    month,
    underemployment_rate_sa,
    underutilisation_rate_sa
  ) |>
  pivot_longer(
    cols = c(
      underemployment_rate_sa,
      underutilisation_rate_sa
    ),
    names_to = "measure",
    values_to = "rate"
  )

ggplot(
  rates_long,
  aes(
    x = month,
    y = rate,
    colour = measure
  )
) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Underemployment and Labour Underutilisation in Australia",
    x = "Year",
    y = "Rate (%)"
  ) +
  theme_minimal()



```
### Do Males and Females Contribute Differently to Underemployment?

At first glance, @fig-gender-share may look a little uneventful—and that is exactly what makes it interesting. While overall underemployment rose and fell dramatically over the decade, the *split between males and females barely changed*.

Male workers consistently accounted for a slightly larger share of total underemployment, contributing around 52–54% of underemployed workers, while females accounted for approximately 46–48%. Even during the COVID-19 disruption, when underemployment surged across Australia, the gender balance remained remarkably stable.

This suggests that both males and females were affected by changing labour market conditions in broadly similar ways. Rather than one gender driving changes in underemployment, the evidence points to economy-wide factors affecting both groups simultaneously.

Most importantly, the difference between the two groups is relatively small compared with the large swings in overall underemployment observed in @fig-underutilisation. This hints that while gender differences certainly exist, they may not be the main reason labour underutilisation changes over time.

So, if males consistently contributed a slightly larger share of underemployment, the next question becomes: did this difference actually matter? To answer that, the next section examines whether the gender composition of underemployment was related to labour underutilisation using correlation and regression analysis.

```{r Deriving-Gender-Variables}
#| label: fig-gender-share
#| fig-cap: "Gender share of total underemployment in Australia, 2016–2025."
labour_data <- labour_data |>
  mutate(
    Tot_male_underemp =
      male_prefers_more_hours +
      male_hours_reduced,

    Tot_female_underemp =
      female_prefers_more_hours +
      female_hours_reduced,

    total_underemp =
      Tot_male_underemp +
      Tot_female_underemp,

    male_share =
      Tot_male_underemp /
      total_underemp * 100,

    female_share =
      Tot_female_underemp /
      total_underemp * 100
  )

share_yearly <- labour_data |>
  mutate(year = year(month)) |>
  group_by(year) |>
  summarise(
    male_share = mean(male_share),
    female_share = mean(female_share)
  ) |>
  pivot_longer(
    cols = c(male_share, female_share),
    names_to = "gender",
    values_to = "share"
  )

share_yearly |>
  ggplot(aes(x = factor(year), y = share, colour = gender, group = gender)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Gender Share of Total Underemployment Over Time",
    x = "Year",
    y = "Share"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  theme_minimal()



```

### What Story Does Correlation tells us?

@fig-gender-imbalance puts the research question to the test by examining whether gender differences in underemployment are linked to labour underutilisation. Surprisingly, the answer appears to be "not really".

The scatterplot shows no clear pattern, and the regression line is almost flat. While males consistently accounted for a slightly larger share of underemployment, these differences did not translate into meaningful changes in labour underutilisation. The regression results support this finding, with a non-significant p-value (0.58) and an R² of just 0.003.

In short, gender differences existed, but they were too small and too stable to explain fluctuations in labour underutilisation. Instead, broader economic events, particularly COVID-19, appear to have been the real drivers of changes in labour market spare capacity.

```{r}
#| label: fig-gender-imbalance
#| fig-cap: "Relationship between gender imbalance in underemployment and labour underutilisation."
#| echo: false
#| results: false
#Creating a Gender Gap

labour_data <- labour_data |>
  mutate(
    share_gap =
      male_share -
      female_share
  )

cor(
  labour_data$share_gap,
  labour_data$underutilisation_rate_sa,
  use = "complete.obs"
)

model <- lm(
  underutilisation_rate_sa ~ share_gap,
  data = labour_data
)

summary(model)

ggplot(
  labour_data,
  aes(
    x = share_gap,
    y = underutilisation_rate_sa
  )
) +
  geom_point(alpha = 0.6) +
  geom_smooth(
    method = "lm",
    se = TRUE
  ) +
  labs(
    title = "Gender Imbalance and Labour Underutilisation",
    x = "Difference in Gender Share (%)",
    y = "Underutilisation Rate (%)"
  ) +
  theme_minimal()

```
```{r Saving-Processed-Dataset}
write_csv(labour_data, "data/labour_data_processed.csv")
write_csv(share_yearly, "data/share_yearly_processed.csv")
```

### So, Does Gender Matter? 

In conclusion, the question is if men and women are getting affected differently? Not as much as we might expect. Is it a good thing? As an economy, this gives clarity of the actual problem. 

While males consistently accounted for a slightly larger share of underemployment, the difference barely changed over the decade and had little impact on labour underutilisation. Instead, the biggest swings in Australia's labour market were driven by broader economic events, most notably the COVID-19 pandemic.

The real story is that underemployment itself remains a major source of unused labour capacity. In other words, Australia's labour market challenge is less about who is underemployed and more about how many workers want more hours than they can get. The findings suggest that improving job quality and access to working hours may be more important than focusing on gender differences alone.

### The Cherry On The Cake! 

Do the current mechanism at place tell the whole labour market story? The ABS says not quite. The [U-series framework](https://www.abs.gov.au/articles/underemployment-and-underutilisation-introducing-u-series) has been introduced which will be fully implemented in 2027, looks beyond people without jobs and shines a spotlight on those who have work but not enough of it, along with others who are only loosely connected to the labour force. 

By widening the lens, the U-series reveals that Australia's spare labour capacity is much larger than unemployment figures alone suggest. In short, the labour market is not just about who has a job and who does not—it's also about who wants more work, more hours, or a better opportunity than the one they currently have.

### References

* Australian Bureau of Statistics (Reference period: February 2025) 'Underemployed Workers' [data set](https://www.abs.gov.au/statistics/labour/employment-and-unemployment/underemployed-workers/feb-2025), Underemployed Workers, Australia, February 2025


* Australian Bureau of Statistics (Reference period: April 2025) 'Underemployment and Underutilisation: Introducing the U-series' [article](https://www.abs.gov.au/articles/underemployment-and-underutilisation-introducing-u-series), Underemployment and Underutilisation: Introducing the U-series, 

* Australian Bureau of Statistics (Reference period: Latest release) 'Labour Force Australia – Underemployment' [data set](https://www.abs.gov.au/statistics/labour/employment-and-unemployment/labour-force-australia/latest-release#underemployment), Labour Force, Australia, 

* Australian Bureau of Statistics (Reference period: 2023) 'Underutilised Labour' [methodology](https://www.abs.gov.au/statistics/detailed-methodology-information/concepts-sources-methods/labour-statistics-concepts-sources-and-methods/2023/concepts-and-sources/underutilised-labour), Labour Statistics: Concepts, Sources and Methods, 2023, 


* R Core Team (2025) R: A Language and Environment for Statistical Computing [software], R Foundation for Statistical Computing, Vienna, Austria, https://www.r-project.org/

* Wickham H. (2016) ggplot2: Elegant Graphics for Data Analysis [R package](https://ggplot2.tidyverse.org/), Springer-Verlag New York, 

* Grolemund G. and Wickham H. (2011) lubridate: Make Dealing with Dates a Little Easier [R package](https://lubridate.tidyverse.org/)

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

*Tell us about parts of your data processing or analysis that weren’t “sexy” and wouldn’t typically be included in a blog post. (e.g. Was their any data drudgery or time intensive wrangling? Were there any repetitive tasks or manual tasks? If it was easy, describe what made it easy?*

The initial approach did not exactly fit well to show my skills I have learnt in this unit, hence keeping same dataset which were two initially and then adding more dataset was a bit tedious and the issues while analysing relating to it: 

* Finding the data that was going to answer my questions were multiple, and the metrics also differed. It was difficult to boil down to the right ones. 

* While the trend % was present with seasonal adjustment for COVID-19 years, the actual trends had no data, so I had to be careful that they did not cause an outlier that would completely throw my question out of context. 


*Were there any challenges that you faced in conducting this analysis. These may take the form of data limitations or coding challenges? (e.g. Was there anything in your analysis that you were not anticipating when you started? Did you have to change your intended scope? Did you need to master a new skill? Were there any problems you were proud of solving?)*

* The data such as Underutilisation rate is not present as Gender specific on ABS website, hence I had to manually derive certain variable. 

* The surprise came when my code started to break right after reading my files. THEY WERE NOT READ AS CSV! I had to clean every data with code to make it read like a CSV. 

* My left_join kept breaking and it took a lot of time to understand what might have been wrong in that. 

* Repetitive layers of filtration and tidy process, along with deciding to choose the right kind of visualisation was a lengthy process. 


*Tell us about any imperfect parts of your work and how you would like to expand or improve this analysis in future? Be clear about any limitations or aspects of your analysis that fell beyond scope.*

I think where I had to certainly compromise on my analysis was the actual reflection of COVID 19. In my understanding, the female section would have been affected more, but surprisingly that was very minimal given the limitation of data. I would have also wanted to include payscale, job vacancies and maybe explore the age factor as well here. 

*Tell us about how you used AI in your approach to your assignment. (e.g. Was AI consistently reliable? Did you provide AI with additional context? Was there anywhere where you questioned or verified the AI outputs? Was there anywhere we AI helped you learn something new?)*

AI was helpful to understand about my coding issues. However, it did misguide me to some extent to think my approach did not make sense. It gave me suggestions that did not align with what I wanted. However, it gave me conceptual understanding that my variables made meaning. It helped me understand the linkage and that further helped me critical think how I could use it to use it in the analysis. 

*Submit 4 earlier versions of your .qmd files to show your iterative process. These versions must show iterations of your work (not just a linear progression of Task 1, Task 2, Task 3, citations). Reflect on the progress made between each versions by providing a short statement for markers of what you fixed/learnt/improved/changed between each file. Also like to reflect on what changed between your first checklist submission. (Note: There is an expectation for students familiar with GitHub that you will submit a Github repository and refer to 4 individual commits.)* 

I have used the Github to **Tag** my versions of work. In most cases my issues were to go back and update the steps of data processing. Making changes in wordings and visualisation. Please find the [GitHub](https://github.com/somya090997/ETC5512_Assig4)


# AI Acknowledgement

Generative Artificial Intelligence (ChatGPT by OpenAI) was used to assist with aspects of this assignment. AI support was used for brainstorming potential analytical approaches, understanding ABS labour market concepts, debugging R code, interpreting data cleaning issues, refining visualisation choices, and improving the clarity and structure of written explanations. 

AI-generated suggestions were critically evaluated, modified where necessary, and independently implemented by the author. All data collection, cleaning, analysis, visualisation, interpretation of results, and final conclusions were completed and verified by the author. AI was used as a learning and productivity aid rather than a substitute for original analytical work.

Please find the [AI Prompts](https://chatgpt.com/share/6a2e1945-f8a4-83ec-aa09-9d30349b6813)

:::
